---
title: "OU Benchmark Figures"
author: "Filippo Monti"
format:
  html:
    toc: true
    df-print: paged
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
options(stringsAsFactors = FALSE)

required_packages <- c(
  "dplyr", "forcats", "fs", "ggplot2", "patchwork", "readr",
  "scales", "stringr", "tibble", "tidyr"
)
missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]
if (length(missing_packages) > 0L) {
  stop(
    "Install missing packages before rendering: ",
    paste(missing_packages, collapse = ", "),
    call. = FALSE
  )
}

suppressPackageStartupMessages({
  lapply(required_packages, library, character.only = TRUE)
})

input_file <- knitr::current_input()
input_dir <- if (is.null(input_file)) {
  getwd()
} else {
  dirname(normalizePath(input_file, mustWork = TRUE))
}

benchmark_root <- fs::path_norm(fs::path(input_dir, "..", "..", ".."))
if (!fs::dir_exists(fs::path(benchmark_root, "benchmarks")) ||
    !fs::dir_exists(fs::path(benchmark_root, "code"))) {
  stop("Could not locate scalableOuBenchmarks root from ", input_dir, call. = FALSE)
}
workspace_root <- fs::path_dir(benchmark_root)
benchmark_data_path <- fs::path(benchmark_root, "benchmarks")
benchmark_figure_path <- fs::path(benchmark_root, "figures")
paper_figure_path <- Sys.getenv(
  "OU_PAPER_FIGS",
  fs::path(workspace_root, "stableMatricesOU", "figures")
)

fs::dir_create(c(benchmark_figure_path, paper_figure_path))

read_benchmark_csv <- function(filename) {
  readr::read_csv(fs::path(benchmark_data_path, filename), show_col_types = FALSE)
}

save_benchmark_plot <- function(plot, filename, width, height) {
  paths <- c(
    fs::path(benchmark_figure_path, filename),
    fs::path(paper_figure_path, filename)
  )
  for (path in paths) {
    ggplot2::ggsave(path, plot = plot, width = width, height = height, dpi = 300)
  }
  invisible(paths)
}

save_benchmark_pair <- function(plot, basename, width, height) {
  save_benchmark_plot(plot, paste0(basename, ".pdf"), width, height)
  save_benchmark_plot(plot, paste0(basename, ".png"), width, height)
}

paper_theme <- function(base_size = 9) {
  theme_classic(base_size = base_size, base_family = "serif") +
    theme(
      axis.line = element_line(linewidth = 0.3),
      axis.ticks = element_line(linewidth = 0.3),
      axis.text = element_text(color = "black"),
      legend.position = "bottom",
      legend.title = element_blank(),
      panel.border = element_rect(color = "black", fill = NA, linewidth = 0.35),
      strip.background = element_rect(fill = "grey93", color = "black", linewidth = 0.3),
      strip.text = element_text(face = "bold"),
      plot.title = element_text(face = "bold", hjust = 0),
      plot.subtitle = element_text(color = "grey25")
    )
}

kernel_colors <- c(
  "SMBP + rotation" = "#2F67B1",
  "SMBP block" = "#2F67B1",
  "Dense" = "#D4960A",
  "Known Schur" = "#9B4356",
  "Schur Parlett" = "#D67C2F"
)

kernel_alpha <- c(
  "SMBP + rotation" = 1,
  "SMBP block" = 0.6,
  "Dense" = 1,
  "Known Schur" = 1,
  "Schur Parlett" = 1
)

kernel_linetypes <- c(
  "SMBP + rotation" = "solid",
  "SMBP block" = "solid",
  "Dense" = "longdash",
  "Known Schur" = "longdash",
  "Schur Parlett" = "longdash"
)
```

# Julia Kernel Benchmarks

The kernel panels read the precomputed CSV files in `benchmarks/`. The current
figure filters those files to `K = 4, 8, 16, 32, 64`; no Julia benchmarks are run
by this document.

```{r kernel-data}
schur_smbp <- read_benchmark_csv("schur_vs_smbp_exp_adjoint_results.csv")
edge_push <- read_benchmark_csv("edge_message_push_results.csv")
lyapunov <- read_benchmark_csv("lyapunov_stationary_variance_results.csv")
branch_lyapunov <- read_benchmark_csv("branch_likelihood_with_lyapunov_results.csv")

keep_k <- c(4L, 8L, 16L, 32L, 64L)

kernel_timings <- bind_rows(
  schur_smbp %>%
    filter(K %in% keep_k) %>%
    transmute(
      K,
      benchmark = "Matrix exponential",
      Dense = dense_exp_ms,
      `Known Schur` = schur_exp_ms,
      `Schur Parlett` = schur_parlett_exp_ms,
      `SMBP block` = smbp_block_exp_ms,
      `SMBP + rotation` = smbp_exp_with_rotation_ms
    ),
  schur_smbp %>%
    filter(K %in% keep_k) %>%
    transmute(
      K,
      benchmark = "Frechet adjoint",
      Dense = dense_adj_ms,
      `Known Schur` = schur_adj_ms,
      `SMBP block` = smbp_adj_apply_ms,
      `SMBP + rotation` = smbp_adj_with_rotation_ms
    ),
  lyapunov %>%
    filter(K %in% keep_k) %>%
    transmute(
      K,
      benchmark = "Stationary covariance",
      Dense = dense_lyap_ms,
      `Known Schur` = schur_known_lyap_ms,
      `SMBP + rotation` = smbp_rotation_lyap_ms
    ),
  edge_push %>%
    filter(K %in% keep_k) %>%
    transmute(
      K,
      benchmark = "Full edge message",
      Dense = dense_full_edge_ms,
      `Known Schur` = schur_known_basis_full_edge_ms,
      `SMBP block` = smbp_block_full_edge_ms,
      `SMBP + rotation` = smbp_full_edge_with_rotation_ms
    ),
  branch_lyapunov %>%
    filter(K %in% keep_k) %>%
    transmute(
      K,
      benchmark = "Branch + Lyapunov",
      Dense = dense_branch_lyap_ms,
      `Known Schur` = schur_branch_lyap_ms,
      `SMBP + rotation` = smbp_branch_lyap_rotation_ms
    )
) %>%
  pivot_longer(
    cols = -c(K, benchmark),
    names_to = "method",
    values_to = "milliseconds",
    values_drop_na = TRUE
  ) %>%
  filter(milliseconds > 0) %>%
  mutate(
    method = factor(method, levels = names(kernel_colors)),
    benchmark = factor(
      benchmark,
      levels = c(
        "Matrix exponential", "Frechet adjoint",
        "Stationary covariance", "Full edge message",
        "Branch + Lyapunov"
      )
    )
  )
```

```{r kernel-figure, fig.width=9.2, fig.height=5.4}
kernel_fig <- ggplot(
  kernel_timings,
  aes(
    x = K,
    y = milliseconds,
    color = method,
    alpha = method,
    linetype = method,
    group = method
  )
) +
  geom_line(linewidth = 0.55) +
  geom_point(size = 1.8) +
  facet_wrap(~ benchmark, scales = "free_y", ncol = 3) +
  scale_x_continuous(
    trans = "log2",
    breaks = keep_k,
    labels = keep_k,
    minor_breaks = NULL
  ) +
  scale_y_log10(
    labels = scales::label_scientific(digits = 1),
    minor_breaks = NULL
  ) +
  scale_color_manual(values = kernel_colors, drop = TRUE) +
  scale_alpha_manual(values = kernel_alpha, drop = TRUE) +
  scale_linetype_manual(values = kernel_linetypes, drop = TRUE) +
  labs(
    x = "State dimension K",
    y = "Median time per call (ms)",
    title = "OU kernel microbenchmarks"
  ) +
  paper_theme()

kernel_fig
```

```{r kernel-save}
save_benchmark_pair(kernel_fig, "ou_kernel_benchmarks", width = 9.2, height = 5.4)
```

# Time-Series Benchmarks

These timings are the historical BEAST-vs-Stan summaries recorded in
`benchmarking.md`. They should be replaced by regenerated CSV output before the
final paper figures are locked.

```{r time-series-data}
time_series_summary <- read_benchmark_csv("time_series_summary.csv")

beast_stan_summary <- time_series_summary %>%
  filter(comparison == "beast_stan") %>%
  mutate(
    scenario = factor(scenario, levels = unique(scenario)),
    operation = factor(operation, levels = c("Log likelihood", "Gradient")),
    method = factor(method, levels = c("Stan", "BEAST"))
  )

beast_cache_summary <- time_series_summary %>%
  filter(comparison == "beast_grid") %>%
  mutate(
    scenario = factor(grid, levels = c("Uniform grid", "Irregular grid")),
    operation = factor(operation, levels = c("Log likelihood", "Gradient"))
  )
```

```{r time-series-figure, fig.width=8.4, fig.height=4}
method_colors <- c("Stan" = "#7b5f95", "BEAST" = "#2f6f73")
grid_colors <- c("Uniform grid" = "#2f6f73", "Irregular grid" = "#b15c3b")

beast_stan_fig <- ggplot(
  beast_stan_summary,
  aes(x = operation, y = milliseconds, color = method, shape = method)
) +
  geom_point(position = position_dodge(width = 0.45), size = 2.6) +
  facet_wrap(~ scenario, nrow = 1) +
  scale_y_log10(labels = scales::label_number()) +
  scale_color_manual(values = method_colors) +
  labs(x = NULL, y = "Median time (ms)", title = "BEAST vs Stan") +
  paper_theme() +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))

cache_fig <- ggplot(
  beast_cache_summary,
  aes(x = operation, y = milliseconds, color = scenario, shape = scenario)
) +
  geom_point(position = position_dodge(width = 0.45), size = 2.6) +
  scale_y_log10(labels = scales::label_number()) +
  scale_color_manual(values = grid_colors) +
  labs(x = NULL, y = NULL, title = "BEAST grid cache") +
  paper_theme() +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))

time_series_fig <- beast_stan_fig + cache_fig + plot_layout(widths = c(1.35, 1))
time_series_fig
```

```{r time-series-save}
save_benchmark_pair(time_series_fig, "ou_time_series_benchmarks", width = 8.4, height = 4)
```

# Tree Benchmarks

The tree benchmark CSVs currently contain a smoke-test grid (`K = 2`, `N = 8`).
The plotting code is set up to expand automatically once larger tree runs are
written to the same `results_*.csv` files.

```{r tree-data}
tree_files <- fs::dir_ls(
  fs::path(benchmark_data_path, "benchmarksTrees"),
  regexp = "results_.*[.]csv$"
)

tree_timings <- bind_rows(lapply(tree_files, readr::read_csv, show_col_types = FALSE)) %>%
  filter(status == "ok", median_sec > 0) %>%
  mutate(
    method_label = case_when(
      package %in% c("PCMBase", "PCMBaseCpp") &
        mode == "dense_PCMLik" ~ paste(package, "dense likelihood"),
      package %in% c("PCMBase", "PCMBaseCpp") &
        mode == "schur_PCMLik" ~ paste(package, "Schur likelihood"),
      package == "mvMORPH" ~ "mvMORPH fixed OU",
      package == "mvSLOUCH" ~ "mvSLOUCH fit proxy",
      package == "PCMFit" ~ "PCMFit fit proxy",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(method_label)) %>%
  group_by(package, K, N, method_label) %>%
  summarize(median_sec = median(median_sec), .groups = "drop") %>%
  mutate(
    grid_label = paste0("K = ", K, ", N = ", N),
    method_label = forcats::fct_reorder(method_label, median_sec)
  )
```

```{r tree-figure, fig.width=7.2, fig.height=4.2}
tree_fig <- ggplot(
  tree_timings,
  aes(x = median_sec, y = method_label, color = package, shape = package)
) +
  geom_point(size = 2.6) +
  facet_wrap(~ grid_label, scales = "free_y") +
  scale_x_log10(labels = scales::label_number()) +
  scale_color_manual(
    values = c(
      "PCMBase" = "#0072B2",
      "PCMBaseCpp" = "#56B4E9",
      "mvMORPH" = "#009E73",
      "mvSLOUCH" = "#E69F00",
      "PCMFit" = "#CC79A7"
    )
  ) +
  labs(
    x = "Median time (s)",
    y = NULL,
    title = "Tree OU package benchmarks"
  ) +
  paper_theme() +
  theme(legend.position = "none")

tree_fig
```

```{r tree-save}
save_benchmark_pair(tree_fig, "ou_tree_benchmarks", width = 7.2, height = 4.2)
```

# Output

Figures are written to both:

- `r benchmark_figure_path`
- `r paper_figure_path`
